%%
%% fall-lecture17.tex
%% 
%% Made by Alex Nelson <pqnelson@gmail.com>
%% Login   <alex@lisp>
%% 
%% Started on  2025-11-06T09:54:27-0800
%% Last update 2025-11-06T09:54:27-0800
%% 

\lecture{}

% This was actually at the end of lecture 16
\begin{theorem}[Chevalley, Matsumara 6E]\label{thm:120a:chevalley}
Let $A$ be a Noetherian ring. Let $A$ be an algebra over $A$ of finite
type (i.e., $B=A[x_{1},\dots,x_{n}]/I$ for some ideal $I$).
Let $\phi\colon A\to B$ be the canonical morphism. Let $X=\Spec(A)$
and $Y=\Spec(B)$. Let $f=\preimage\phi\colon Y\to X$.

Then $f(Y')$ of a constructible set $Y'$ in $Y$ is constructible in $X$.
\end{theorem}

Note: localizations are not ``of finite type''.

\begin{proof}
Let us show it only for $Y'=Y$. Let $X_{0}$ be some irreducible closed
subset of $X$. (We will use Proposition~\ref{prop:fall-lec16:prop-6c}.)
Any irreducible closed subset of some $\Spec$ of a ring may be written
as
\begin{equation}
X_{0}=V(P_{0}),
\end{equation}
where $P_{0}\in\Spec(A)$. We have 
\begin{equation}
A'=A/P_{0},\quad\mbox{and}\quad B'=B/P_{0}B.
\end{equation}
Suppose that $X_{0}\cap f(Y)$ is dense. Conceptually we have something
like
\begin{equation}
\begin{array}{rcl}
  \Spec(B')=f^{-1}(X_{0}) & \to & X_{0}=\Spec(A')\\
  \rotatebox{-90}{$\subset$} & & \rotatebox{-90}{$\subset$}\\
  Y & \xrightarrow{f} & X
\end{array}
\end{equation}
Then the map
\begin{equation}
\phi'\colon A'\to B'
\end{equation}
is injective by Lemma~\ref{lm:fall-lec16:matsumara-6-2}. Replacing
$A$, $B$, $\phi$ by $A'$, $B'$, $\phi'$ (respectively) gives us the
following claim to prove:
\begin{quote}
If $A$ is a Noetherian domain and $B$ is a ring containing $A$ (since
$\phi$ is injective) and $B$ is finitely-generated over $A$, then
there exists a nonzero $a\in A$ such that the elementary open set
$D(a)$ of $X=\Spec(A)$ is contained in $f(Y)$ when $Y=\Spec(B)$ and
$f$ is the canonical map.
\end{quote}
So by going to the quotient rings simplifies the claim, to the point
when we can prove it by pure algebra. (Recall $D(a)=\{P\in\Spec(A)\mid a\notin P\}$.)
Now we can write $B=A[x_{1},\dots,x_{n}]$. Suppose $x_{1}$, \dots,
$x_{r}$ are algebraically independent, while $x_{r+1}$, \dots, $x_{n}$
satisfy some algebraic relations over
$A^{*}=A[x_{1},\dots,x_{r}]$. Choose some relations of the form
\begin{equation}
g_{j,0}(x)x_{j}^{d_{j}}+g_{j,1}(x)x_{j}^{d_{j}-1}+\dots=0,
\end{equation}
where the coefficients $g_{j,i}(x)\in A^{*}$ and $g_{j,i}\neq0$. The
trick is to multiply the leading coefficients together
$\prod^{n}_{j=r+1}g_{j,0}(x)$ which is a nonzero polynomial in
$A[x_{1},\dots, x_{r}]$. Pick $a$ to be any coefficient of
\begin{equation}
g(x) = \prod^{n}_{j=r+1}g_{j,0}(x).
\end{equation}
We claim: if $a\notin P_{0}$, then $P_{0}\in\Im(f)$. (Hence
$D(a)\subset f(Y)$.) Suppose we have some $P\in\Spec(A)$ such that $a\notin P$.
Then let $P^{*}=PA^{*}=P[x_{1},\dots,x_{r}]$. Then $\prod g_{j,0}\notin P^{*}$.
So $B_{P^{*}}$ is integral over $A^{*}_{P^{*}}$.
(Recall when we had $A\to B$ and $B$ is integral over $A$ and $A$ is a
Noetherian domain, then this satisfies lying over\dots)
Then there exists some prime ideal $P_{1}$ of $B_{P^{*}}$ lying over $P^{*}A^{*}_{P^{*}}$.
We have
\begin{equation}
P_{1}\cap A=P_{1}\cap A^{*}\cap A=P[x_{1},\dots,x_{r}]\cap A=P^{*},
\end{equation}
which implies
\begin{equation}
P=P_{1}\cap A=(P_{1}\cap B)\cap A\in f(\Spec(B)).
\end{equation}
Hence the result.
\end{proof}

\begin{remark}
Why do we care about algebras ``of finite type''? Well, they are where
our geometric intuition (that maximal ideals are points, and prime
ideals are curves) really works. So we can draw these diagrams ``honestly''.
\end{remark}